A thermodynamic view of dusty protoplanetary disks

### 2017-05-09 summary.tex
misc edits to the main text overall

As noted in \S\ref{loc_iso_eos}, dust settling causes 
$\widetilde{T}$ to \emph{increase} away from the midplane. This contrasts
to previous studies of GI in vertically stratified disks
 where the temperature decreases
from the midplane \citep[e.g.][]{mamat10, kim12,lin14c}. 
While we expect only the total surface density and
characteristic temperatures are relevant to stability  
\citep{toomre64}, a non-trivial vertical temperature
structure, induced by dust, may modify the vertical structure of 3D
waves and unstable modes. We will examine this problem in a following 
study.

Another potential connection to previous results for gas disks is  
gravito-turbulence. Cooling, self-gravitating gaseous disks 
sustain a turbulent state where shock heating due to gravitational 
instabilities are balanced by thermal losses
\citep{gammie01}. Now, dust-gas drift behaves like thermal diffusion
or cooling (Eq. \ref{eff_energy}), similar to how radiative cooling 
appears in models of gravito-turbulent disks. Thus it may be
conceivable that a massive dusty disk admit `dusty
gravito-turbulence'.

@@

The dusty/adiabatic gas analogy imply the condition for RWI in a
polytropic dusty gas disk is an extremum   
\begin{align}
  \mathcal{V} =
  \frac{\kappa^2}{2\Omega\Sigma}\frac{1}{K^{2/\Gamma}}\left(1+
    \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}}\right)^2.    
\end{align}
Thus we expect that RWI may also be triggered by extrema 
in the dust-to-gas ratio, e.g. narrow dust rings/gaps. This may lead
to direct formation of dusty vortices, as opposed to dust-trapping by  
a pre-existing gas vortex \citep{barge95,lyra13}.

@@

Isothermal disks have $\tcool=0$, so the pure gas ConO cannot
exist. However, we have shown that dust-gas drag provides an effective
energy source/sink to the mixture (\S\ref{energy_analogy}). It would be
interesting to explore whether or not dust-gas drag can provide the 
required `cooling' to enable a `dusty convective 
overstability'.

@@

The `zombie vortex instability' (ZVI) is a non-axisymmetric,
non-linear instability discovered in pure gas disks
\citep{marcus15,umurhan16d}. It operates by having finite-amplitude perturbations
exciting `critical layers' which roll 
up into vortices, exciting further critical layers. The process
repeats itself and the result is an array of vortices. These critical 
layers are locations where the intrinsic wave frequency of a 
disturbance matches the local vertical buoyancy frequency 
\citep{marcus13}.

Since the necessary physical ingredient for the ZVI is vertical
buoyancy, it requires an almost purely adiabatic gas ($\tcool\OmK\gg
1$). This limits its applicability in typical protoplanetary disks to 
$\lesssim 1$AU \citep{lesur16, malygin17}. However, this consideration
assumes a dust-free disk.

On the other hand, we have show that dust-loading an isothermal disk
also induces an effective buoyancy. % This effect is maximized when dust is
% perfectly coupled to the gas
. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading, and thus produce `dusty ZVI'. This may allow the zombie
vortices to exist at $\gtrsim 1$AU in PPDs.

@@

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. We will apply 
our modified code --- d\textsc{Pluto} --- to the streaming instability
and disk-planet interactions in an upcoming publication (Lin et al.,
in preparation).

@@

We obtain an analog of the Solberg-Hoiland criteria for
axisymmetric stability for strictly isothermal, perfectly-coupled dusty gas.  
Applying this to typical protoplanetary disks, we find that 
the vertical shear associated with dust 
 layers \emph{cannot} lead to axisymmetric  
  instabilities, however thin the dust layer is.    
Instead, sharp radial  edges in the dust-to-gas ratio could destabilized the
disk, as these imply sharp gradients in the disk's effective entropy
profile. Alternatively, if the dust is vertically well-mixed, then any
radial gradient in $\rhod/\rhog$ can destabilize the disk.

@@

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We
explicitly show in SI the evolution of gas
pressure lags behind dust density; and is in fact a general property of 
overstabilities driven by strong dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies there exists a time interval where 
the gas pressure of a parcel of the dusty-gas mixture is increasing whilst dust is already being expelled.  
The dusty gas then does positive work that amplifies 
oscillations. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.

@@

We also discuss future applications of our thermodynamic framework to 
study dusty protoplanetary disks. Because isothermal dusty gas has an
effective entropy, we suggest that purely hydrodynamic processes, such
as the Rossby Wave Instability \citep{li00} or the `Zombie Vortex
Instability' \citep{marcus15} could have dusty counter-parts.   
Furthermore, hydrodynamic instabilities driven by thermal cooling, such as VSI in
stably-stratified disks \citep{lin15} or the `Convective
Overstability' \citep{klahr14, lyra14},  may also find dusty analogs 
because finite dust-gas drag is equivalent to a heat sink/source.

>>>>

As noted in \S\ref{loc_iso_eos}, dust settling causes 
$\widetilde{T}$ to \emph{increase} away from the midplane. This contrasts
to previous studies of GI in vertically stratified disks
 where the temperature decreases
from the midplane \citep[e.g.][]{mamat10, kim12,lin14c}. 
While we expect only the total surface density and
characteristic temperatures are relevant to stability  
\citep{toomre64}, a non-trivial vertical temperature
structure, induced by dust, may modify the vertical structure of 3D
waves and unstable modes.

Another potential connection to previous results for gas disks is  
gravito-turbulence. Cooling, self-gravitating gaseous disks 
sustain a turbulent state where shock heating due to gravitational 
instabilities are balanced by radiative cooling 
\citep{gammie01}. Since dust-gas drift appears as a diffusion or 
cooling in our framework (Eq. \ref{eff_energy}), it may be conceivable
to have `dusty gravito-turbulence'. As self-gravity increases the
local density, the associated pressure maxima attract dust-particles,
but the back-reaction onto the gas may try to flatten the pressure
bump \citep{taki16}.

@@

The dusty/adiabatic gas analogy imply the condition for RWI in a
polytropic dusty gas disk is an extremum   
\begin{align}
  \mathcal{V} =
  \frac{\kappa^2}{2\Omega\Sigma}\frac{1}{K^{2/\Gamma}}\left(1+
    \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}}\right)^2.    
\end{align}
Thus RWI may also be triggered by extrema 
in the dust-to-gas ratio, e.g. narrow dust rings/gaps. This may lead 
to direct formation of dusty vortices, as opposed to dust-trapping by  
a pre-existing gas vortex \citep{barge95,lyra13}.

@@

Isothermal disks have $\tcool=0$, so the pure gas ConO cannot
exist. However, we have shown that dust-gas drag provides an effective
energy source/sink to the mixture (\S\ref{energy_analogy}). It would be
interesting to explore whether or not dust-gas drag can play the role
of cooling to enable a `dusty convective 
overstability'.

@@

The `zombie vortex instability' (ZVI) is a non-axisymmetric,
non-linear instability discovered in pure gas disks
\citep{marcus15,umurhan16d}. It operates by having finite-amplitude perturbations
exciting `critical layers',  where the intrinsic wave frequency of the
perturbation matches the local vertical buoyancy frequency 
\citep{marcus13}. These critical layers roll 
up into vortices, exciting further critical layers. The process
repeats itself and the result is an array of vortices.

Since the necessary physical ingredient for the ZVI is vertical
buoyancy, it requires an almost purely adiabatic gas ($\tcool\OmK\gg
1$). This limits its applicability in typical protoplanetary disks to 
$\lesssim 1$AU \citep{lesur16, malygin17}. However, this consideration
assumes a dust-free disk as far as the dynamics is concerned.

On the other hand, we have show that dust-loading induces an effective
buoyancy even in isothermal disks. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading, and thus produce `dusty ZVI'. This may allow the zombie
vortices to develop $\gtrsim 1$AU in PPDs.

@@

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code \citep{mignone07} into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. We will apply 
our modified code --- d\textsc{Pluto} --- to the streaming instability
and disk-planet interactions in an upcoming publication (Lin et al.,
in preparation).

@@

We obtain the equivalent Solberg-Hoiland criteria for the 
axisymmetric stability of strictly isothermal, perfectly-coupled dusty gas.  
Applying this to typical protoplanetary disks, we find that 
the vertical shear associated with dust 
 layers \emph{cannot} lead to axisymmetric  
  instabilities, however thin the dust layer is.    
Instead, sharp radial  edges in the dust-to-gas ratio could destabilized the
disk, as these imply sharp gradients in the disk's effective entropy
profile. Alternatively, if the dust is vertically well-mixed, then any
radial gradient in $\rhod/\rhog$ can destabilize the disk.

@@

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We
explicitly show in SI the evolution of gas
pressure lags behind dust density. This is in fact a general property of 
instabilities driven by dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies there exists a time interval where 
the gas pressure of a parcel of the dusty-gas mixture is increasing whilst dust is already being expelled.  
The dusty gas then does positive work that amplifies 
oscillations. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.

@@

We also discuss future applications of our thermodynamic framework to 
study dusty protoplanetary disks. Because isothermal dusty gas has an
effective entropy, we suggest that purely hydrodynamic processes, such
as the Rossby Wave Instability \citep{li00} or the `Zombie Vortex
Instability' \citep{marcus15}, where entropy plays a role, could have dusty counter-parts.   
Furthermore, hydrodynamic instabilities driven by thermal cooling, such as VSI in
stably-stratified disks \citep{lin15} or the `Convective
Overstability' \citep{klahr14, lyra14},  may also find dusty analogs 
because finite dust-gas drag is equivalent to a heat sink/source.
